Extended Abstract

### 2021-05-04 paper/paper.tex
Add type system

\newcommand{\picalc}{$\pi$-calculus}
\newcommand{\Picalc}{$\pi$-Calculus}

@@

\textbf{TODO: recursion/replication} \textbf{TODO: bounded linear
types?} \textbf{TODO: type assertions}

Consider a \(\pi\)-calculus with usage annotations on channels. These
usage annotations are one of \(0\), \(1\), or \(\omega\). \(0\) must not
be used, \(1\) must be used exactly once, \(\omega\) can be used as many
times as desired. Sum and products contain usage annotations on their
leafs.

@@

\subsection{Types}\label{types}

Let \(\gamma\) be a context of metavariable kinds, where a kind is
either a \texttt{type} or a \texttt{usage}. Then \(\gamma \ni k\) is a
choice of a \(k\) inside \(\gamma\). \[
\begin{aligned}
S, T : \texttt{Type}~\gamma &= \texttt{mvar}~(\gamma \ni \texttt{type}) ~|~ \#_{i,o}t ~|~ \top ~|~ S + T ~|~ S \times T \\
i,o : \texttt{Usage}~\gamma &= \texttt{mvar}~(\gamma \ni \texttt{usage}) ~|~ 0\cdot ~|~ 1\cdot ~|~ \omega\cdot
\end{aligned}
\]

The type system is the usual standard type system based on context
splits that one defines for the pi calculus. TODO: insert type system
definition here:.

Usage annotations are \(0\cdot\), \(1\cdot\), \(\omega\cdot\). Context
splits are defined as follows:

\[
\begin{aligned}
\forall x.    && x           &= x + 0\cdot \\
\forall x.    && x           &= 0\cdot + x \\
\forall x ~y. && \omega\cdot &= y + z
\end{aligned}
\]

Only the usage annotations at the upper level of a channel type are
split: the usage annotations of any channel type that is send as data
are preserved on all three sides of the split.

\hypertarget{inference}{%
\section{Inference}\label{inference}}

>>>>

% Typing rules
\usepackage{mathpartir}
\mprset {sep=1em} % Horizontal space between premises

\newcommand{\picalc}{$\pi$-calculus}
\newcommand{\Picalc}{$\pi$-Calculus}
\newcommand{\rulename}[1]{{\tiny \textsc{(#1)}}}
\newcommand{\constr}[1]{\textcolor{olive}{\mathtt{#1}}}
\newcommand{\type}[1]{\textcolor{blue}{\mathtt{#1}}}

% Syntax types
\newcommand{\Fin}[1]{\type{Fin}~#1}
\newcommand{\sExpr}[1]{\type{Expr}~#1}
\newcommand{\sProc}[1]{\type{Proc}~#1}
\newcommand{\tvar}[2]{#1 ~\type{\ni_t}~ #2}
\newcommand{\tkind}[2]{#1 ~\type{\vdash_t}~ #2}
\newcommand{\ttype}[1]{\type{Type}~#1}
\newcommand{\tusage}[1]{\type{Usage}~#1}
\newcommand{\tCtx}[2]{\type{Ctx}_{#1}~#2}
\newcommand{\tSplit}[3]{#1~\type{=}~#2~\type{\uplus}~#3}
\newcommand{\tun}[1]{\type{un}~#1}
\newcommand{\tVar}[3]{#1 ~ \type{\ni} ~ #2 ~ \type{:} ~ #3}
\newcommand{\tExpr}[3]{#1 ~ \type{\vdash} ~ #2 ~ \type{:} ~ #3}
\newcommand{\tProc}[2]{#1 ~ \type{\vdash} ~ #2}

% Syntax constructors
\newcommand{\sunit}{\constr{unit}}
\newcommand{\svar}{\constr{var}}
\newcommand{\sfst}{\constr{fst}}
\newcommand{\ssnd}{\constr{snd}}
\newcommand{\sinl}{\constr{inl}}
\newcommand{\sinr}{\constr{inr}}
\newcommand{\spair}{\constr{pair}}
\newcommand{\send}{\constr{end}}
\newcommand{\snew}{\constr{new}}
\newcommand{\scomp}{\constr{comp}}
\newcommand{\srecv}{\constr{recv}}
\newcommand{\ssend}{\constr{send}}
\newcommand{\scase}{\constr{case}}

% Kind constructors
\newcommand{\ktype}{\constr{ty}}
\newcommand{\kusage}{\constr{us}}

% Type constructors
\newcommand{\tmvar}{\constr{mvar}}
\newcommand{\tchan}{\constr{chan}}
\newcommand{\tunit}{\constr{unit}}
\newcommand{\tsum}{\constr{sum}}
\newcommand{\tprod}{\constr{prod}}
\newcommand{\tzero}{\constr{0\cdot}}
\newcommand{\tone}{\constr{1\cdot}}
\newcommand{\tomega}{\constr{\omega\cdot}}

@@

\section{Syntax}\label{syntax}

Expressions $e_n$ and $f_n$ are of type $\sExpr{n}$, processes $p_n$ and $q_n$ are of type $\sProc{n}$, variable references $i_n$ are of type $\Fin{n}$.
\[
\begin{aligned}[c]
  e_n ~ f_n  :=
  &~ \sunit \\
  |&~ \svar~i_n \\
  |&~ \spair~e_n~f_n \\
  |&~ \sfst~e_n ~|~  \ssnd~e_n \\
  |&~ \sinl~e_n ~|~  \sinr~e_n
\end{aligned}
\begin{aligned}[c]
  p_n ~ q_n  :=
  &~ \send ~|~  \snew~p_{1+n} \\
  |&~ \srecv~e_n~p_{1+n} \\
  |&~ \ssend~e_n~f_n~p_n \\
  |&~ \scomp~p_n~q_n \\
  |&~ \scase~e_n~p_{1+n}~q_{1+n}
\end{aligned}
\]
\todo{add recursion}
\todo{add type assertions}

@@

Both types and usage annotations allow for \emph{holes} or \emph{metavariables}.
We use the same set of metavariables for both types and usage annotations -- this makes their handling more uniform.
We use a context of kinds \(\gamma\) to keep track of whether a metavariable is of type kind $\ktype$ or usage annotation kind $\kusage$: a choice of kind \(k\) in a kinding context \(\gamma\) is denoted as $\tvar{\gamma}{k}$.

We define well kinded types and usage annotations simultaneously so that we are able to provide a uniform unification and context merge algorithm later on (note that we define $\tsum$ and $\tprod$ and $\tzero$, $\tone$ and $\tomega$ in one go to save space):
\begin{mathpar}
  \inferrule {m : \tvar{\gamma}{k}} {\tmvar~m : \tkind{\gamma}{k}}

  \inferrule {
    i : \tkind{\gamma}{\kusage} \\
    o : \tkind{\gamma}{\kusage} \\
    t : \tkind{\gamma}{\ktype}}
  {\tchan~i~o~t : \tkind{\gamma}{\ktype}}

  \inferrule { }
  {\tunit : \tkind{\gamma}{\ktype}}

  \inferrule {s : \tkind{\gamma}{\ktype} \\ t : \tkind{\gamma}{\ktype} }
  {\tsum~s~t ~|~ \tprod~s~t : \tkind{\gamma}{\ktype}}

  \inferrule { } {\tzero ~|~ \tone ~|~ \tomega : \tkind{\gamma}{\kusage}}
\end{mathpar}
We will henceforth use $\ttype{\gamma}$ to stand for $\tkind{\gamma}{\ktype}$ and $\tusage{\gamma}$ to stand for $\tkind{\gamma}{\kusage}$.

A context $\Gamma$ of type $\tCtx{n}{\gamma}$ is a list of $\ttype{\gamma}$ of size $n$.
We define context splits $\tSplit{\Gamma}{\Delta}{\Theta}$ pointwise on types.
Splits on types are defined pointwise recursively on usage annotations -- note however that only the usage annotations on the top level of a channel (that is, not within its payload) are split.
Splits on usage annotations are defined as follows -- note that they are not necessarily unique:
\begin{mathpar}
  \inferrule { } {\tSplit{x}{x}{\tzero}}

  \inferrule { } {\tSplit{x}{\tzero}{x}}

  \inferrule { } {\tSplit{\tomega}{x}{y}}
\end{mathpar}

A usage annotation $x$, type $t$ or context $\Gamma$ is unrestricted (shared, non-linear) $\tun{\Gamma}$ if it can be split into itself: $\tSplit{\Gamma}{\Gamma}{\Gamma}$ (respectively $x$ and $t$) \todo{cite padovani}.
\todo{typing judgment for variables}

We can finally define our typing judgment for expressions and processes. \todo{formation rules double lined}

\begin{mathpar}
  \inferrule { \Gamma : \tCtx{n}{\gamma} \\ e : \sExpr{n} \\ t : \ttype{\gamma}}
             { \tExpr{\Gamma}{e}{t} }

  \inferrule { \tVar{\Gamma}{i}{t} } { \tExpr{\Gamma}{\svar~i}{t} } \rulename{var}

  \inferrule { \tun{\Gamma} } { \tExpr{\Gamma}{\sunit}{\tunit} } \rulename{unit}

  \inferrule { \tExpr{\Gamma}{e}{\tprod~t~s} \\ \tun{s}} { \tExpr{\Gamma}{\sfst~e}{t}} \rulename{fst}

  \inferrule { \tExpr{\Gamma}{e}{t} } { \tExpr{\Gamma}{\sinl~e}{\tsum~t~s}} \rulename{inl}

  \inferrule { \tSplit{\Gamma}{\Delta}{\Theta} \\ \tExpr{\Delta}{e}{s} \\ \tExpr{\Theta}{f}{t} }
             { \tExpr{\Gamma}{\spair~e~f}{\tprod~s~t}}
             \rulename{pair}

  \inferrule { \Gamma : \tCtx{n}{\gamma} \\ p : \sProc{n}}
             { \tProc{\Gamma}{p} }

  \inferrule { \tun{\Gamma} } { \tProc{\Gamma}{\send} } \rulename{end}

  \inferrule { t : \ttype{\gamma} \\ \tProc{\Gamma , t}{p} } { \tProc{\Gamma}{\snew~p} }

  \inferrule { \tSplit{\Gamma}{\Delta}{\Theta} \\ \tExpr{\Delta}{e}{\tchan~\tone~\tzero~t} \\ \tProc{\Theta,t}{p} }
             { \tProc{\Gamma}{\srecv~e~p} }

  \inferrule { \tSplit{\Gamma}{\Delta}{\Theta} \\ \tExpr{\Delta}{e}{\tsum~s~t} \\ \tProc{\Theta,s}{p} \\ \tProc{\Theta,t}{q} }
             { \tProc{\Gamma}{\scase~e~p~q} }
\end{mathpar}

\section{Inference}\label{inference}
